A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-08 podc_paper/algorithm-2.tex
factored out subroutine CompleteDequeue, added some description of GC algorithm details

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
The nodes in the RBT each contain a pointer to a block, as well as the standard RBT node attributes: child pointers and a colour bit; this means that when an update on a RBT uses node-copying, the pointer to the block
gets copied, rather than copying the block itself.)
Each RBT is initialized with an empty block with index 0.

\here{Talk about what happens if a block is not found in an RBT}

We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.

\here{Walk reader through the code (and how it is different from basic versino)}

\op{MaxBlock} is a query operation on a RBT that returns the block with the maximum \fld{index}.
We could keep a pointer to the maximum block in the root of the RBT so that \op{MaxBlock}
can be done in constant time.

\here{Emphasize that RBT nodes have \emph{pointers to} blocks.  Thus when RBT does node copying,
it copies the pointer to the block; it does not make a copy of the block itself.
This is important because if we mark a block that has been node-copied, we want the block to still be marked.  Similarly for super field.}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}$,\\
		$\fld{index}\assign h+1$}\label{enqNewGC}
	}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1$,\\
	    $\fld{index}\assign h+1$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{h+1}, 1, \fa}$
    \gc{
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\If{$\var{res} = \nl$} \linecomment{mark blocks}
		\State $\Call{IndexDequeue}{\var{leaf}, h+1, 1, \tr}$
	\EndIf
	\State \Return{\var{res}}
    }
    \label{deqRestGC}
\EndFunction{Dequeue}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
		}
    \EndIf

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

\gc{
\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
	\State \linecomment{set \fld{super} field of last block in $T$}
	\State \linecomment{and in $v$'s children's RBTs}
	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
	\If{$v\neq \var{root}$}	
		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
	\EndIf
	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
    \EndFor \label{endHelpChild1GC}
\EndFunction{SetSuper}
}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

@@

\caption{Bounded-space queue implementation, continued.\label{pseudocode1GC}}
\end{figure}

@@

\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

>>>>

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
To avoid confusion, we use nodes to refer to the nodes of the ordering tree, and blocks to refer
to the nodes of a RBT (since the RBT stores blocks).

The space-bounded implementation uses an additional shared array so that processes can access one another's leaves to perform helping.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\end{itemize}
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
Each \typ{Node} has an additional array field to help determine which of its blocks are safe to discard.
\begin{itemize}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the node that process $k$ }\\
\mbox{ }\Comment{has accessed, either to find the enqueued item for its dequeue}\\
\mbox{ }\Comment{or during a \nl\ dequeue}
\end{itemize}
The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index 0.
The \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in constant time, without affecting the time of other RBT operations.

\here{Talk about what happens if a block is not found in an RBT}
\here{Talk about \emph{marking} a block to indicate that any blocks with smaller indices will no longer
be needed by operations that haven't started yet.}

Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
The \op{Propagate} and \op{FindResponse} routines are unchanged.
Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
visited by the routines to mark the blocks they visit in those nodes.
Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
separate routine called \op{CompleteDequeue} so that it can also be used by other processes
helping to complete the operation.

Once a dequeue has been propagated to the root, 
\bf{\op CompleteDequeue} computes its answer.  This routine can also be called by helpers 
who want to help complete the dequeue.
In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
it also updates \var{last} entries if the computed response is \nl.
It does this by calling \op{IndexDequeue} again, this time with \tr\ as its final parameter.
This extra parameter has been added to 
\op{IndexDequeue} to indicate whether it should mark blocks that it traverses (which all contain the dequeue).

\here{Discuss new routines AddBlock, SetSuper, Help, and largely modified Refresh}

\here{does RBT have to have pointer to block rather than block itself so that updates to super field work correctly?}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,\\
		$\fld{index}\assign h$}\label{enqNewGC}
	}
    \State \Call{Append}{\var{B}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,\\
	    $\fld{index}\assign h$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \gc{\State \Return \Call{CompleteDequeue}{\var{leaf}, $h$}}
\EndFunction{Dequeue}

\gc{
\Function{Object}{CompleteDequeue}{\var{leaf}, $h$}
    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1, \fa}$
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\If{$\var{res} = \nl$} \linecomment{update \var{last} values}
		\State $\Call{IndexDequeue}{\var{leaf}, h, 1, \tr}$
	\EndIf
	\State \Return{\var{res}}
\EndFunction{CompleteDequeue}
}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{48}
\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
		}
    \EndIf

@@

\spac
\gc{
\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
	\State \linecomment{set \fld{super} field of last block in $T$}
	\State \linecomment{and in $v$'s children's RBTs}
	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
	\If{$v\neq \var{root}$}	
		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
	\EndIf
	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
    \EndFor \label{endHelpChild1GC}
\EndFunction{SetSuper}
}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDequeue}{\ell, \var{B}.\fld{index}}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{48}

@@

\caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
\end{figure}

@@

\end{algorithmic}
\caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }
\end{figure}
